The Development of a Collaborative Tool to Teach Debugging

### 2020-12-07 paper/paper.tex
final paper

A large number of tools exist to facilitate collaborative programming.
COVID19 has greatly increased the demand for tools that not only make
collaboration easier, but make it easy remotely.  The tool that was
most influential in the design of the collaborative debugger,
Replit\cite{replit}, enables remote collaborative programming.

@@

\lstinline{memoization} will be covered in detail in the following
section on the frontend of the collaborative debugger (\ref{webapp}).
All three examples show how teachers can design debugging sessions
around specific concepts.  The collaborative debugger should make it
easy to distribute these sessions and for students to learn from them
collaboratively.
\par

Students using all three examples would benefit greatly from
visualization tools to help understand pointer locations, the process
space, and data structures.  As discussed in \ref{vistools}, adding
visualization tools to the collaborative debugger should be relatively
easy.
\par

@@

Before diving into the construction of the frontend webapp, it may be
helpful to look at the steps two students would take to debug the
\lstinline{memoization} example program remotely.  This example is
covered in more detail in \ref{current}.

@@

Notice that the debugging window in the above example is called
``rr''. rr is ``a lightweight tool for recording, replaying and
debugging execution of applications''\cite{rr-repo}. rr allows a
programmer to record the execution of a program on any compatible
machine and replay the execution later.  This enhances GDB's ability
to ``time-travel'' when debugging, using commands such as
\lstinline{reverse-continue} and
\lstinline{reverse-stepi}\cite{gdbman} to step backwards and forwards
through a program's execution.  Through a novel encapsulation of the
execution space, rr is able to deterministically record and replay the
execution of syscalls and other process behavior that differs
run-to-run.  This is invaluable when trying to debug behavior that is
not entirely dependent on the code being debugged.  A typical workflow
in rr consists of recording an inexplicable error, replaying execution
to find the area in which the error occurs, and then narrowing in on
the bug not by re-running the entire program, but by progressing back
and forth through execution in the problem area.
\par

@@

In debugging the \lstinline{memoization} example program, users will
want to display the call stack from within each implementation of the
nth Fibonacci number function.  To stop in \lstinline{fib_rec}, the
recursive implementation, one user will want to set a breakpoint at
line 52 in the source file.  To do this they will type the command
\lstinline{break 52} and hit enter.  The command will be sent to the
backend, and all users in the same debug session (17734 in Figure
\ref{frontend:rrterm}) will see the debuggers response:\\
\lstinline{Breakpoint 1 at 0x561c8fede303: file memo.c, line 52.}\\
Users who did not send the command \lstinline{c} will see it displayed
along with the response.

@@

The collaborative debugger consists of a distributed Kubernetes
backend and frontend React webapp.  Kubernetes was chosen for the
backend primarily so that a Pod could be created dynamically for each
debugging session.

@@

The Pod configuration schema for most Pods in the collaborative
debugger is fairly generic.  It consists of a \lstinline{name}, an
\lstinline{image} sourced from the container registry, and in the case
of pods that need to interact with a load balancer, an
\lstinline{app}.

@@

For statically created Pods, the \lstinline{kubectl apply} command is
used to create new Pods.  Kubernetes pulls the container image defined
in the schema from the container registry and starts the Pod with any
necessary commands.  The database Pod is connected to a long-term
storage volume at this time.  Upon successful creation, the Pod is
ready to interact with any necessary load balancers.

@@

At the heart of the collaborative debugger is the RR Debug Session.
Every time a user wants to debug a new program execution, an RR Debug
Session Pod is dynamically created by the API server.  The new Pod is
assigned a unique five digit identification number when it is created.
This five digit number is used as the channel for the RR Debug
Session, separating its communication from other RR Debug Session
Pods.  Clients, RR Message Server Pods, and the RR Debug Session Pod
connect through the RR Message Load Balancer.  Clients and RR Debug
Sessions send messages using the channel assigned to the pod at
creation time.  A diagram of one full communication cycle between two
clients and an RR Debug Session Pod is outlined in the diagram below:

@@

Current example pod startup scripts compile the program to be debugged
and record execution when the pod is created.

@@

Effort is taken in the design of the RR Debug Session and RR Message
Server to ensure students receive a pleasant and near-identical
experience to debugging locally in the future.  The system of
inheritance used to create example debug session pods is designed to
be as simple as possible, though a more fully featured frontend could
streamline this process further in the future.

@@

The API server is implemented using Flask.  The server program
consists of handlers for the various API routes and instances of two
classes which communicate with the database/Kubernetes API:
\lstinline{TranslationPodManager} and \lstinline{UserManager}.  The
full structure of the API Server program, \lstinline{api.py} is shown
below:

@@

Care was taken to abstract as much as possible in the design of the
API server and the classes it instantiates.  The two classes
instantiated by \lstinline{api.py}, \lstinline{TranslationPodManager}
and \lstinline{UserManager}, each instantiate another class which
interacts with the database, \lstinline{DatabaseController}, as an
instance variable.  \lstinline{TranslationPodManager} further
instantiates the Kubernetes Core V1 Python API to communicate with the
cluster in order to create/delete pods.  This process of abstraction
ensures that the database, underlying cluster structure, and
authentication methods can all change without significant changes
needing to be made to \lstinline{api.py}.
\par

@@

The Kubernetes API is then finally used to create the Pod.  The Pod is
created in the default namespace.  This could be changed to logically
isolate dynamically and statically created Pods with minimal hassle,
since Kubernetes' DNS resolution can connect pods across namespaces.
If Pod creation fails, an error will be thrown.  Otherwise, a loop is
used to wait until \lstinline{rrtranslation.py} is running inside the
Pod.  Once the Pod is ready, \lstinline{channel} is returned.

@@

The Frontend/API server connects all components of the collaborative
debugger together.  It authenticates users through the database, joins
the frontend and Kubernetes cluster through the dynamic creation of
Pods, and starts the interaction between users and the programs they
wish to debug.  Care has been taken to make it modular and extensible
within reason, and additional authentication or pod creation features
should be easy to add.
\par

@@

The following section details the first few steps students take to
explore the \lstinline{memoization} example using the collaborative
debugger.  Students A and B begin by examining the call stack at its
deepest point in the recursive nth Fibonacci number function.  The
terminal interface to rr and the source code view have been
color-inverted to improve readability in this paper.

To start a new session, Student A selects ``memoization'' from the
list of example sessions.\pagebreak

In order to join Student A's session, Student B uses session code
``71825''.  Meanwhile, Student A is presented with the the main view
of the collaborative debugger.  On the right is a view of the source
code for the \lstinline{memoization} example.  On the left is the
terminal interface to rr.\pagebreak

@@

The most significant addition towards this goal would be that of
frontend visualization tools.  A simple visualization of the stack and
registers would greatly expand students' understanding of the program
being debugged in all three example sessions currently included with
the collaborative debugger.  gdb/mi supports commands which give
structured output of both these aspects of the execution space, and
the frontend has been designed to make adding visualizations of them
simple.  Adding visualization features will bring the collaborative
debugger much closer towards hopefully fulfilling it's second
requirement.

@@

A study similar to those undertaken in \cite{10.1145/1026487.1008043},
\cite{10.1145/1145287.1145293}, and \cite{10.1145/3361721.3361724}
could be done to test both these aspects.  Students participating in
an introductory systems programming class could be split into three
groups.  Group one could study debugging individually, using rr by
themselves to progress through debugging sessions designed by the
instructor.  Group two could study debugging in pairs, relying on
existing tools such as video-conferencing software to facilitate
collaboration.  In the final group, pairs would use the collaborative
debugger to work together on the same debugging sessions as groups one
and two.  Students' individual and pair performance in debugging both
unseen and self-generated programs could then be tested.

@@

The collaborative debugger aims to encourage collaboration and build
students knowledge of debugging through careful design.  Many
components of a distributed system come together to create a platform
that hopefully benefits both teachers and students.  Through
deliberate design, the collaborative debugger can conceivably be the
ideal platform to teach and learn debugging, both remotely and
in-person.

>>>>

A large number of tools exist to facilitate collaborative programming.
COVID-19 has greatly increased the demand for tools that not only make
collaboration easier, but make it easy remotely.  The tool that was
most influential in the design of the collaborative debugger,
Replit\cite{replit}, enables remote collaborative programming.

@@

% \lstinline{memoization} will be covered in detail in the following
% section on the frontend of the collaborative debugger (\ref{webapp}).
% All three examples show how teachers can design debugging sessions
% around specific concepts.
The collaborative debugger should make it easy to distribute these
sessions and for students to learn from them collaboratively.
\par

The following section details the first few steps students take to
explore the \lstinline{memoization} example using the current version
of the collaborative debugger.  This example, which focuses more on
learning about how functions execute than on fixing bugs, highlights
the ways in which the collaborative debugger enables pair programming.
\par

Students A and B begin by examining the call stack at its deepest
point in the recursive nth Fibonacci number function.  The terminal
interface to rr and the source code view have been color-inverted to
improve readability in this paper.

Students A and B both log into the collaborative debugger.  To start a
new session, Student A selects ``memoization'' from the list of
example sessions.\pagebreak

In order to join Student A's session, Student B uses session code
``71825''.  Meanwhile, Student A is presented with the the main view
of the collaborative debugger.  On the left is a view of the source
code for the \lstinline{memoization} example.  On the right is the
terminal interface to rr, the enhancement to gdb used by the
collaborative debugger.  Students debug recordings of program
execution, which can be replayed deterministically. rr is covered in
detail in \ref{rr}.\pagebreak

Though the collaborative debugger makes it easy for students in the
above example to practice using a debugger together, more features
could be added to help improve students' understanding of the program
being debugged. Students using all three examples would benefit
greatly from visualization tools to help understand pointer locations,
the process space, and data structures.
\par

In \lstinline{memoization}, a simple stack visualization that updated
with each command sent to the debugger would allow students to more
easily see changes in the process space.  Rather than having to
repeatedly \lstinline{continue} and \lstinline{backtrace} to visualize
the stack, students could simply step through relevant portions of the
program and watch as the stack updated.  This sort of visualization
pairs particularly well with \lstinline{reverse-continue} or
\lstinline{reverse-step}(\lstinline{i}), allowing students to move back and forth
over particularly significant stack changes.  Visualization tools also
aid students' understanding of debugging---students can see how typing
\lstinline{bt} to backtrace or \lstinline{info frame} to view
information about the current stack frame provide the same information
as the visualization tools they are already familiar with.  As
discussed in \ref{vistools}, adding visualization tools to the
collaborative debugger should be relatively easy.
\par

@@

An overview of the \lstinline{memoization} example discussed in
\ref{current} is below.  This pattern is followed by all debugging
sessions in the frontend of the collaborative debugger.

@@

Notice that the debugging window in the above example is called
``rr''. rr is ``a lightweight tool for recording, replaying and
debugging execution of applications''\cite{rr-repo}. rr allows a
programmer to record the execution of a program on any compatible
machine and replay the execution later.  This enhances GDB's ability
to ``time-travel'' when debugging, using commands such as
\lstinline{reverse-continue} and
\lstinline{reverse-step}(\lstinline{i})\cite{gdbman} to step backwards
and forwards through a program's execution.  Through a novel
encapsulation of the execution space, rr is able to deterministically
record and replay the execution of syscalls and other process behavior
that differs run-to-run.  This is invaluable when trying to debug
behavior that is not entirely dependent on the code being debugged.  A
typical workflow in rr consists of recording an inexplicable error,
replaying execution to find the area in which the error occurs, and
then narrowing in on the bug not by re-running the entire program, but
by progressing back and forth through execution in the problem area.
\par

@@

In debugging the \lstinline{memoization} example program, users will
want to display the call stack from within each implementation of the
nth Fibonacci number function.  To stop in \lstinline{fib_rec}, the
recursive implementation, one user will want to set a breakpoint at
line 52 in the source file.  To do this they will type the command
\lstinline{break 52} and hit enter.  The command will be sent to the
backend, and all users in the same debug session (17734 in Figure
\ref{frontend:rrterm}) will see the debuggers response:\\
\lstinline{Breakpoint 1 at 0x561c8fede303: file memo.c, line 52.}\\
Users who did not send the command \lstinline{break 52} will see it
displayed along with the response.

@@

The collaborative debugger consists of a distributed Kubernetes
backend and frontend React webapp.  Kubernetes was chosen for the
backend primarily so that a Pod could be created dynamically for each
debugging session. \pagebreak

@@

The Pod configuration schema for most Pods in the collaborative
debugger is fairly generic.  It consists of a \lstinline{name}, an
\lstinline{image} sourced from the container registry, and in the case
of Pods that need to interact with a load balancer, an
\lstinline{app}.

@@

For statically created Pods, the \lstinline{kubectl apply} command is
used to create new Pods.  Kubernetes pulls the container image defined
in the schema from the container registry and starts the Pod with any
necessary commands.  The Database Pod is connected to a long-term
storage volume at this time.  Upon successful creation, the Pod is
ready to interact with any necessary load balancers.

@@

At the heart of the collaborative debugger is the RR Debug Session.
Every time a user wants to debug a new program execution, an RR Debug
Session Pod is dynamically created by the API server.  The new Pod is
assigned a unique five digit identification number when it is created.
This five digit number is used as the channel for the RR Debug
Session, separating its communication from other RR Debug Session
Pods.  Clients, RR Message Server Pods, and the RR Debug Session Pod
connect through the RR Message Load Balancer.  Clients and RR Debug
Sessions send messages using the channel assigned to the Pod at
creation time.  A diagram of one full communication cycle between two
clients and an RR Debug Session Pod is outlined in the diagram below:

@@

Current example Pod startup scripts compile the program to be debugged
and record execution when the Pod is created.

@@

Effort is taken in the design of the RR Debug Session and RR Message
Server to ensure students receive a pleasant and near-identical
experience to debugging locally in the future.  The system of
inheritance used to create example Debug Session Pods is designed to
be as simple as possible, though a more fully featured frontend could
streamline this process further in the future.

@@

The API server is implemented using Flask.  The server program
consists of handlers for the various API routes and instances of two
classes which communicate with the database and/or Kubernetes API:
\lstinline{TranslationPodManager} and \lstinline{UserManager}.  The
full structure of the API Server program, \lstinline{api.py} is shown
below:

@@

Care was taken to abstract as much as possible in the design of the
API server and the classes it instantiates.  The two classes
instantiated by \lstinline{api.py}, \lstinline{TranslationPodManager}
and \lstinline{UserManager}, each instantiate another class which
interacts with the database, \lstinline{DatabaseController}, as an
instance variable.  \lstinline{TranslationPodManager} further
instantiates the Kubernetes Core V1 Python API to communicate with the
cluster in order to create/delete Pods.  This process of abstraction
ensures that the database, underlying cluster structure, and
authentication methods can all change without significant changes
needing to be made to \lstinline{api.py}.
\par

@@

The Kubernetes API is then finally used to create the Pod.  The Pod is
created in the default namespace.  This could be changed to logically
isolate dynamically and statically created Pods with minimal hassle,
since Kubernetes' DNS resolution can connect Pods across namespaces.
If Pod creation fails, an error will be thrown.  Otherwise, a loop is
used to wait until \lstinline{rrtranslation.py} is running inside the
Pod.  Once the Pod is ready, \lstinline{channel} is returned.

@@

The Frontend/API server connects all components of the collaborative
debugger together.  It authenticates users through the database, joins
the frontend and Kubernetes cluster through the dynamic creation of
Pods, and starts the interaction between users and the programs they
wish to debug.  Care has been taken to make it modular and extensible
within reason, and additional authentication or Pod creation features
should be easy to add.
\par

@@

@@

The most significant addition towards this goal would be that of
frontend visualization tools.  A simple visualization of the stack and
registers would greatly expand students' understanding of the program
being debugged in all three example sessions currently included with
the collaborative debugger.  gdb/mi supports commands which give
structured output of both these aspects of the execution space, and
the frontend has been designed to make adding visualizations of them
simple.  Adding visualization features will bring the collaborative
debugger much closer towards hopefully fulfilling its second
requirement.

@@

A study similar to those undertaken in \cite{10.1145/1026487.1008043},
\cite{10.1145/1145287.1145293}, and \cite{10.1145/3361721.3361724}
could be done to test both of these aspects.  Students participating
in an introductory systems programming class could be split into three
groups.  Group one could study debugging individually, using rr by
themselves to progress through debugging sessions designed by the
instructor.  Group two could study debugging in pairs, relying on
existing tools such as video-conferencing software to facilitate
collaboration.  In the final group, pairs would use the collaborative
debugger to work together on the same debugging sessions as groups one
and two.  Students' individual and pair performance in debugging both
unseen and self-generated programs could then be tested.

@@

The collaborative debugger aims to encourage collaboration and build
students' knowledge of debugging through careful design.  Many
components of a distributed system come together to create a platform
that hopefully benefits both teachers and students.  Through
deliberate design, the collaborative debugger can conceivably be the
ideal platform to teach and learn debugging, both remotely and
in-person.
